---
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F, eval = T, results = 'hide',
                      fig.align = 'center', fig.width = 8, fig.height = 8)
knitr::opts_knit$set(root.dir = here::here())
```
```{r}
library(tidyverse)
library(ggridges)
library(viridis)
library(DT)
```
```{r}
#++++++++++++++++++++++++++++++++++++++++++
### Part 0: Import Data and Setup     ####
#++++++++++++++++++++++++++++++++++++++++++
# sim data from 01-polySimIBD_data/05-run_sims.R
simdat <- readRDS("data/sim_data/goDISC_simulated_gendata.RDS") %>%
  dplyr::mutate(modname = paste(modnameNe, "rep", rep, sep = "-")) %>%
  dplyr::select(c("modname", "IBDcalc"))
# location data from 01-polySimIBD_data/02-cartesian_empiric_liftover.R
locatdat <- readRDS("data/sim_data/sim_params/locatcombo.rds")
# discent runs search-grid
search_discdat <- readRDS("data/sim_data/goDISC_search-DISC_simulated_gendata.RDS") %>% 
  dplyr::mutate( overallmodname = stringr::str_extract(string = modname, pattern = "^(.*?)(?=-rep)") )
# discent runs full
discdat <- readRDS("data/sim_data/goDISC_full-DISC_simulated_gendata.RDS")
# bring together
sim_disc_dat <- dplyr::left_join(simdat, discdat, by = "modname")

```


## DISC Results: Overview of Search Parameters and Cost

### Parameter Search
Given concerns for a non-convex loss surface, we performed a grid search over the below starting parameter values. Searches were performed for 10,000 steps. 

| Parameter               | Values Tested                       | Notes                                      |
|-------------------------|-------------------------------------|--------------------------------------------|
| Inbreeding Coefficient (f) | 0.01, 0.05, 0.1, 0.25, 0.33          | Same start value applied across all 25 demes       |
| Migration Rate (M)         | 0.5, 5, 10, 50, 100                 | -      |
| Learning Rate (α)          | 1e-4, 1e-3, 1e-2, 1e-1              | ADAM optimizer step size                    |
| Regularization (λ)         | 1e-6, 1e-5, 1e-4, 1e-3              | L2 penalty on migration rate                |

- We used fixed ADAM hyperparameters (default)


### Cost Summary
```{r}
search_discdat %>% 
  dplyr::group_by(overallmodname) %>% 
  dplyr::summarise(
    mincost = min(cost),
    firstqcost = quantile(cost, prob = 0.25),
    mediancost = quantile(cost, prob = 0.5),
    meancost = mean(cost),
    firstqcost = quantile(cost, prob = 0.75),
    sdcost = sd(cost),
    maxcost = max(cost) ) %>% 
  mutate(across(where(is.numeric), ~ round(., digits = 0))) %>% 
  DT::datatable(.,
                rownames = F,
                extensions='Buttons',
                options = list(
                  searching = T,
                  pageLength = 10,
                  dom = 'Bfrtip',
                  autoWidth = TRUE,
                  buttons = c('csv')))
```

### Visualization of **Migration-Start & F-Start** versus Cost
```{r}
search_discdat %>% 
  dplyr::mutate(fm = paste(paste0("m: ", mstart), paste0("f: ", fstart), sep = "; ")) %>% 
  ggplot() + 
  ggridges::geom_density_ridges(aes(x = cost, y = factor(fm), fill = overallmodname), color = NA, alpha = 0.25, scale = 0.8) +
  labs(x = "Cost", y = "F-M Start") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle(label = "Overall Cost by Migration Surface for F & M Start Params")


search_discdat %>% 
  dplyr::mutate(fm = paste(paste0("m: ", mstart), paste0("f: ", fstart), sep = "; ")) %>%
  dplyr::filter(cost < 30000) %>% 
  ggplot() + 
  ggridges::geom_density_ridges(aes(x = cost, y = factor(fm), fill = overallmodname), color = NA, alpha = 0.25, scale = 0.8) +
  labs(x = "Cost", y = "F-M Start") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle(label = "Overall Cost by Migration Surface for F & M Start Params",
          subtitle = "Right-Truncuated for Outliers")


```

### Visualization of **Learning Rate** versus Cost
```{r}
search_discdat %>% 
  ggplot() + 
  ggridges::geom_density_ridges(aes(x = cost, y = factor(learningrate), fill = overallmodname), color = NA, alpha = 0.25, scale = 0.8) +
  labs(x = "Cost", y = "Learning Rate Start") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle(label = "Overall Cost by Migration Surface for Learning Rate")


search_discdat %>% 
  dplyr::filter(cost < 1e5) %>% 
  ggplot() + 
  ggridges::geom_density_ridges(aes(x = cost, y = factor(learningrate), fill = overallmodname), color = NA, alpha = 0.25, scale = 0.8) +
  labs(x = "Cost", y = "Learning Rate Start") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle(label = "Overall Cost by Migration Surface for Learning Rate",
          subtitle = "Right-Truncuated for Outliers")


```


### Visualization of **Lambda** versus Cost
```{r}
search_discdat %>% 
  ggplot() + 
  ggridges::geom_density_ridges(aes(x = cost, y = factor(lambda), fill = overallmodname), color = NA, alpha = 0.25, scale = 0.8) +
  labs(x = "Cost", y = "Lambda Start") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle(label = "Overall Cost by Migration Surface for Lambda")


search_discdat %>% 
  dplyr::filter(cost < 25000) %>% 
  ggplot() + 
  ggridges::geom_density_ridges(aes(x = cost, y = factor(lambda), fill = overallmodname), color = NA, alpha = 0.25, scale = 0.8) +
  labs(x = "Cost", y = "Lambda Start") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle(label = "Overall Cost by Migration Surface for Lambda",
          subtitle = "Right-Truncuated for Outliers")


```

### Summary of Cost versus **Starting Parameters**
```{r}
# data
d1 <- search_discdat %>% 
  dplyr::mutate(llr = paste(paste0("LR: ", learningrate), paste0("L: ", lambda)), sep = "; ",
                fm = paste(paste0("m: ", mstart), paste0("f: ", fstart), sep = "; "),
                llrfm = paste(llr, fm, sep = "; ")) 
# plot
d1 %>% 
  ggplot() + 
  ggridges::geom_density_ridges(aes(x = cost, y = factor(llrfm), fill = overallmodname), color = NA, alpha = 0.25, scale = 0.8) +
  labs(x = "Cost", y = "Lambda; Learning Rate; F; M Start") +
  theme(
    panel.background = element_blank(),       # remove gray bg
    plot.background = element_blank(),        # remove outer bg
    panel.grid.major = element_blank(),       # remove major gridlines
    panel.grid.minor = element_blank(),       # remove minor gridlines
    axis.line = element_line(color = "black"), # keep axis lines
    axis.text.x = element_text(angle = 90),
    axis.text.y = element_text(size = 1)) 




d1 %>% 
  dplyr::group_by(llrfm) %>% 
  dplyr::summarise(
    mincost = min(cost),
    firstqcost = quantile(cost, prob = 0.25),
    mediancost = quantile(cost, prob = 0.5),
    meancost = mean(cost),
    firstqcost = quantile(cost, prob = 0.75),
    sdcost = sd(cost),
    maxcost = max(cost) ) %>% 
  mutate(across(where(is.numeric), ~ round(., digits = 0))) %>% 
  DT::datatable(.,
                rownames = F,
                extensions='Buttons',
                options = list(
                  searching = T,
                  pageLength = 10,
                  dom = 'Bfrtip',
                  autoWidth = TRUE,
                  buttons = c('csv')))


```


### Summary of Cost versus **Starting Parameters** and **Model**
```{r}
# data
d1 <- search_discdat %>% 
  dplyr::mutate(llr = paste(paste0("LR: ", learningrate), paste0("L: ", lambda)), sep = "; ",
                fm = paste(paste0("m: ", mstart), paste0("f: ", fstart), sep = "; "),
                llrfm = paste(llr, fm, sep = "; "),
                modllrfm = paste(overallmodname, llrfm, sep = "; ")) 


d1 %>% 
  dplyr::group_by(modllrfm) %>% 
  dplyr::summarise(
    mincost = min(cost),
    firstqcost = quantile(cost, prob = 0.25),
    mediancost = quantile(cost, prob = 0.5),
    meancost = mean(cost),
    firstqcost = quantile(cost, prob = 0.75),
    sdcost = sd(cost),
    maxcost = max(cost) ) %>% 
  mutate(across(where(is.numeric), ~ round(., digits = 0))) %>% 
  DT::datatable(.,
                rownames = F,
                extensions='Buttons',
                options = list(
                  searching = T,
                  pageLength = 10,
                  dom = 'Bfrtip',
                  autoWidth = TRUE,
                  buttons = c('csv')))


```


### Summary of Cost versus **Starting Parameters** and **Model** and **Rep**
```{r}

# random rep 
search_discdat %>% 
  dplyr::mutate(llr = paste(paste0("LR: ", learningrate), paste0("L: ", lambda), sep = ";") ,
                fm = paste(paste0("m: ", mstart), paste0("f: ", fstart), sep = "; "),
                llrfm = paste(llr, fm, sep = "; ")) %>% 
  dplyr::filter(modname == "bound-rep-10") %>% 
  ggplot() + 
  geom_point(aes(x = fm, y = llr, color = cost)) +
  scale_color_viridis_c() +
  theme(
    panel.background = element_blank(),       # remove gray bg
    plot.background = element_blank(),        # remove outer bg
    panel.grid.major = element_blank(),       # remove major gridlines
    panel.grid.minor = element_blank(),       # remove minor gridlines
    axis.line = element_line(color = "black"), # keep axis lines
    axis.text.x = element_text(angle = 90)) +
  ggtitle("Random Selection of Model: Bound, Rep 10")




# plot
search_discdat %>% 
  dplyr::mutate(llr = paste(paste0("LR: ", learningrate), paste0("L: ", lambda), sep = ";") ,
                fm = paste(paste0("m: ", mstart), paste0("f: ", fstart), sep = "; "),
                llrfm = paste(llr, fm, sep = "; ")) %>% 
  dplyr::filter(modname == "bound-rep-6") %>% 
  ggplot() + 
  geom_point(aes(x = fm, y = llr, color = cost)) +
  scale_color_viridis_c() +
  theme(
    panel.background = element_blank(),       # remove gray bg
    plot.background = element_blank(),        # remove outer bg
    panel.grid.major = element_blank(),       # remove major gridlines
    panel.grid.minor = element_blank(),       # remove minor gridlines
    axis.line = element_line(color = "black"), # keep axis lines
    axis.text.x = element_text(angle = 90)) +
  ggtitle("Random Selection of Model: Bound, Rep 6")



# random rep 
search_discdat %>% 
  dplyr::mutate(llr = paste(paste0("LR: ", learningrate), paste0("L: ", lambda), sep = ";") ,
                fm = paste(paste0("m: ", mstart), paste0("f: ", fstart), sep = "; "),
                llrfm = paste(llr, fm, sep = "; ")) %>% 
  dplyr::filter(modname == "torus-rep-10") %>% 
  ggplot() + 
  geom_point(aes(x = fm, y = llr, color = cost)) +
  scale_color_viridis_c() +
  theme(
    panel.background = element_blank(),       # remove gray bg
    plot.background = element_blank(),        # remove outer bg
    panel.grid.major = element_blank(),       # remove major gridlines
    panel.grid.minor = element_blank(),       # remove minor gridlines
    axis.line = element_line(color = "black"), # keep axis lines
    axis.text.x = element_text(angle = 90)) +
  ggtitle("Random Selection of Model: Torus, Rep 10")




# plot
search_discdat %>% 
  dplyr::mutate(llr = paste(paste0("LR: ", learningrate), paste0("L: ", lambda), sep = ";") ,
                fm = paste(paste0("m: ", mstart), paste0("f: ", fstart), sep = "; "),
                llrfm = paste(llr, fm, sep = "; ")) %>% 
  dplyr::filter(modname == "torus-rep-6") %>% 
  ggplot() + 
  geom_point(aes(x = fm, y = llr, color = cost)) +
  scale_color_viridis_c() +
  theme(
    panel.background = element_blank(),       # remove gray bg
    plot.background = element_blank(),        # remove outer bg
    panel.grid.major = element_blank(),       # remove major gridlines
    panel.grid.minor = element_blank(),       # remove minor gridlines
    axis.line = element_line(color = "black"), # keep axis lines
    axis.text.x = element_text(angle = 90)) +
  ggtitle("Random Selection of Model: Torus, Rep 6")





```



### Summary of Cost versus **Final Fs** and **Final Ms**
```{r}

# random rep 
search_discdat %>% 
  dplyr::filter(modname == "bound-rep-10") %>% 
  dplyr::mutate(meanFi = purrr::map_dbl(mod, function(x){mean(x$Final_Fis)}),
                finalm = purrr::map_dbl(mod, function(x){x$Final_m})) %>% 
  ggplot() + 
  geom_point(aes(x = meanFi, y = finalm, color = cost)) +
  scale_color_viridis_c() +
  theme(
    panel.background = element_blank(),       # remove gray bg
    plot.background = element_blank(),        # remove outer bg
    panel.grid.major = element_blank(),       # remove major gridlines
    panel.grid.minor = element_blank(),       # remove minor gridlines
    axis.line = element_line(color = "black"), # keep axis lines
    axis.text.x = element_text(angle = 90)) +
  ggtitle("Random Selection of Model: Bound, Rep 10")




# plot
search_discdat %>% 
  dplyr::filter(modname == "bound-rep-6") %>% 
  dplyr::mutate(meanFi = purrr::map_dbl(mod, function(x){mean(x$Final_Fis)}),
                finalm = purrr::map_dbl(mod, function(x){x$Final_m})) %>% 
  ggplot() + 
  geom_point(aes(x = meanFi, y = finalm, color = cost)) +
  scale_color_viridis_c() +
  theme(
    panel.background = element_blank(),       # remove gray bg
    plot.background = element_blank(),        # remove outer bg
    panel.grid.major = element_blank(),       # remove major gridlines
    panel.grid.minor = element_blank(),       # remove minor gridlines
    axis.line = element_line(color = "black"), # keep axis lines
    axis.text.x = element_text(angle = 90)) +
  ggtitle("Random Selection of Model: Bound, Rep 6")





```

