---
title: ""
author: "Nicholas F. Brazeau"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    highlight: textmate
    theme: lumen
    toc: yes
    toc_float: yes
    toc_depth: 3
    code_folding: hide
editor_options: 
  chunk_output_type: console
---


```{r, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F, eval = T, results = 'hide',
                      fig.align = 'center', fig.width = 8, fig.height = 8)
knitr::opts_knit$set(root.dir = here::here())
```
```{r}
library(tidyverse)
remotes::install_github("nickbrazeau/discent")
library(discent)
library(cowplot)
library(viridis)
library(stringr)
library(DT)
set.seed(48)
```


```{r}
#++++++++++++++++++++++++++++++++++++++++++
### Part 0: Import Data and Setup     ####
#++++++++++++++++++++++++++++++++++++++++++
# sim data from 01-polySimIBD_data/05-run_sims.R
simdat <- readRDS("data/sim_data/goDISC_simulated_gendata.RDS") %>%
  dplyr::mutate(modname = paste(modnameNe, "rep", rep, sep = "-")) %>%
  dplyr::select(c("modname", "IBDcalc"))
# location data from 01-polySimIBD_data/02-cartesian_empiric_liftover.R
locatdat <- readRDS("data/sim_data/sim_params/locatcombo.rds")
# discent runs full
discdat <- readRDS("data/sim_data/goDISC_full-DISC_simulated_gendata.RDS") %>% 
  dplyr::rename(mod = cost)

#............................................................
# Helper Functions
#...........................................................
plot_swf_sim <- function(locatdat, IBDcalc, mod, modname, threshold, alpha = 0.5){
  #......................
  # tidy up location data for Sim IBD Plot
  #......................
  locatdat1 <- locatdat %>%
    dplyr::select(dplyr::contains("1")) %>%
    dplyr::filter(!duplicated(.))
  locatdat2 <- locatdat %>%
    dplyr::select(dplyr::contains("2")) %>%
    dplyr::filter(!duplicated(.))
  # bring together
  simplotdat <- IBDcalc %>%
    dplyr::left_join(., y = locatdat1, by = "deme1") %>%
    dplyr::left_join(., y = locatdat2, by = "deme2")
  #......................
  # plot sim IBD
  #......................
  p1 <- simplotdat %>%
    dplyr::filter(gendist > threshold) %>%
    ggplot() +
    geom_point(aes(x = deme1longnum, y = deme1latnum),
               color = "#d9d9d9", alpha = 0.2) +
    geom_segment(aes(x = deme1longnum, y = deme1latnum,
                     xend = deme2longnum, yend = deme2latnum,
                     color = gendist), alpha = alpha) +
    viridis::scale_color_viridis("IBD Dist.") +
    theme_minimal() +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.text = element_blank(),
          axis.title = element_blank(),
          axis.ticks = element_blank())

  #......................
  # mod data clean up
  #......................
  discplotdat <- dplyr::bind_cols(mod$deme_key, mod$Final_Fis)
  colnames(discplotdat) <- c("deme1", "key", "disc")
  p2 <- discplotdat %>%
    dplyr::left_join(.,locatdat1, by = "deme1") %>%
    ggplot() +
    geom_point(aes(x = deme1longnum, y = deme1latnum, fill = disc),
               shape = 21,
               size = 3) +
    theme_minimal() +
    scale_fill_viridis("DISC", option = "mako") +
    theme(axis.title = element_blank())

  #......................
  # add title and subtitle
  #......................
  # Combine plots side by side
  plots <- cowplot::plot_grid(p1, p2, labels = NULL)

  # Create title and subtitle as separate ggdraws
  titlechar <- sim_disc_dat
  title <- ggdraw() +
    draw_label(modname,
               fontface = 'bold', size = 18, x = 0.5, hjust = 0.5)

  subtitle1 <- ggdraw() +
    draw_label(paste("IBD cutoff for viz: ", threshold),
               size = 12, x = 0.5, hjust = 0.5)
  subtitle2 <- ggdraw() +
    draw_label(paste("Final M calculated:", round(mod$Final_m, digits = 3)),
               size = 12, x = 0.5, hjust = 0.5)
  subtitle <- cowplot::plot_grid(subtitle1, subtitle2, labels = NULL)


  # Combine title, subtitle, and plots
  final_plot <- plot_grid(
    title,
    plots,
    subtitle,
    ncol = 1,
    rel_heights = c(0.1,1,0.05)
  )

  # Display
  final_plot
}



#++++++++++++++++++++++++++++++++++++++++++
### Part 1: Visualize Results     ####
#++++++++++++++++++++++++++++++++++++++++++
sim_disc_dat <- dplyr::left_join(simdat, discdat, by = "modname")

sim_disc_dat$viz <- purrr::pmap(sim_disc_dat[,c("IBDcalc", "mod", "modname")], plot_swf_sim,
                                threshold = 0.1, alpha = 0.5, locatdat = locatdat)


```

```{r}
# tidy up viz data
torus <- sim_disc_dat %>% 
  dplyr::filter(stringr::str_detect(modname, "torus-rep-*[0-9]"))
bound <- sim_disc_dat %>% 
  dplyr::filter(stringr::str_detect(modname, "bound-rep-*[0-9]"))
dexter <- sim_disc_dat %>% 
  dplyr::filter(stringr::str_detect(modname, "dexter-rep-*[0-9]"))
torusNe <- sim_disc_dat %>% 
  dplyr::filter(stringr::str_detect(modname, "torus-NeVary-rep-*[0-9]"))
boundNe <- sim_disc_dat %>% 
  dplyr::filter(stringr::str_detect(modname, "bound-NeVary-rep-*[0-9]"))
```

## Isotropic Homogenous with Periodic Boundaries {.tabset}

### Rep 1
```{r} 
torus$viz[[1]]
``` 

### Rep 2
```{r} 
torus$viz[[2]]
``` 

### Rep 3
```{r} 
torus$viz[[3]]
``` 

### Rep 4
```{r} 
torus$viz[[4]]
``` 

### Rep 5
```{r} 
torus$viz[[5]]
``` 

### Rep 6
```{r} 
torus$viz[[6]]
``` 

### Rep 7
```{r} 
torus$viz[[7]]
``` 

### Rep 8
```{r} 
torus$viz[[8]]
``` 

### Rep 9
```{r} 
torus$viz[[9]]
``` 

### Rep 10
```{r} 
torus$viz[[10]]
``` 


## Isotropic Homogenous with Reflecting Boundaries {.tabset}

### Rep 1
```{r} 
bound$viz[[1]]
``` 

### Rep 2
```{r} 
bound$viz[[2]]
``` 

### Rep 3
```{r} 
bound$viz[[3]]
``` 

### Rep 4
```{r} 
bound$viz[[4]]
``` 

### Rep 5
```{r} 
bound$viz[[5]]
``` 

### Rep 6
```{r} 
bound$viz[[6]]
``` 

### Rep 7
```{r} 
bound$viz[[7]]
``` 

### Rep 8
```{r} 
bound$viz[[8]]
``` 

### Rep 9
```{r} 
bound$viz[[9]]
``` 

### Rep 10
```{r} 
bound$viz[[10]]
``` 


## Anisotropic (no left) Homogenous with Reflecting Boundaries {.tabset}

### Rep 1
```{r} 
dexter$viz[[1]]
``` 

### Rep 2
```{r} 
dexter$viz[[2]]
``` 

### Rep 3
```{r} 
dexter$viz[[3]]
``` 

### Rep 4
```{r} 
dexter$viz[[4]]
``` 

### Rep 5
```{r} 
dexter$viz[[5]]
``` 

### Rep 6
```{r} 
dexter$viz[[6]]
``` 

### Rep 7
```{r} 
dexter$viz[[7]]
``` 

### Rep 8
```{r} 
dexter$viz[[8]]
``` 

### Rep 9
```{r} 
dexter$viz[[9]]
``` 

### Rep 10
```{r} 
dexter$viz[[10]]
``` 


## Isotropic Homogenous with Periodic Boundaries and Varying Ne {.tabset}

### Rep 1
```{r} 
torusNe$viz[[1]]
``` 

### Rep 2
```{r} 
torusNe$viz[[2]]
``` 

### Rep 3
```{r} 
torusNe$viz[[3]]
``` 

### Rep 4
```{r} 
torusNe$viz[[4]]
``` 

### Rep 5
```{r} 
torusNe$viz[[5]]
``` 

### Rep 6
```{r} 
torusNe$viz[[6]]
``` 

### Rep 7
```{r} 
torusNe$viz[[7]]
``` 

### Rep 8
```{r} 
torusNe$viz[[8]]
``` 

### Rep 9
```{r} 
torusNe$viz[[9]]
``` 

### Rep 10
```{r} 
torusNe$viz[[10]]
``` 


## Isotropic Homogenous with Reflecting Boundaries and Varying Ne {.tabset}

### Rep 1
```{r} 
boundNe$viz[[1]]
``` 

### Rep 2
```{r} 
boundNe$viz[[2]]
``` 

### Rep 3
```{r} 
boundNe$viz[[3]]
``` 

### Rep 4
```{r} 
boundNe$viz[[4]]
``` 

### Rep 5
```{r} 
boundNe$viz[[5]]
``` 

### Rep 6
```{r} 
boundNe$viz[[6]]
``` 

### Rep 7
```{r} 
boundNe$viz[[7]]
``` 

### Rep 8
```{r} 
boundNe$viz[[8]]
``` 

### Rep 9
```{r} 
boundNe$viz[[9]]
``` 

### Rep 10
```{r} 
boundNe$viz[[10]]
``` 
