---
title: "Evaluating DISCent Model"
author: "Nicholas F. Brazeau"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    highlight: textmate
    theme: lumen
    toc: yes
    toc_float: yes
    toc_depth: 3
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = T, 
                      warning = F, 
                      message = F, 
                      eval = T, 
                      results = 'hide',
                      fig.align = 'center', 
                      fig.width = 9, fig.height = 7)
knitr::opts_knit$set(root.dir = here::here(""))
```
```{r}
library(tidyverse)
# remotes::install_github("nickbrazeau/discent", ref = "develop")
library(discent)
source("R/utils.R")
```



# Overview 
**Model conceptualization**: 
$$
r_{ij} = \left(\frac{f_i + f_j}{2}\right) exp\left\{ \frac{-d_{ij}}{M} \right\}
$$


**Coding Decisions**: 
1. M and F parameters have different learning rates (uncoupled) 
2. M must be positive 
2. M is normalized as part of input 
3. ADAM gradient descent optimizer


## Simulated Data 
Using `polySimIBD` have five frameworks of simulation, which for each we ran 100 independent realizations (i.e. same migration, same Ne = different seed). For each framework, there are 25 demes laid out on a 5x5 square plane. Frameworks are as follows: 

1. Isolation by Distance, Ne Constant (migrations are $\frac{1}{dist}$)    
2. Lattice (can only move to neighbors, no diagonal moves; bounded by borders), Ne Constant   
3. Torus (as above, but no longitudinal boundaries = cylinder), Ne Constant  
4. IBDist, Ne: 5,25,50,100,250 across rows of 5x5 square matrix 
4. Isolation by Distance with wrong distances (to analyze cost)
\br
**tlim here now 25 generations** 
\br

## Go DISCent Table for Starting Parameters and Learning Rates
```
fstartsvec <- c(0.1, 0.3, 0.5)
mstartsvec <- c(10, 25, 50, 100, 250) # standardizing geodistance
flearnsvec <- mlearnsvec <-  c(1e-2, 1e-3, 1e-4)
search_grid <- expand.grid(fstartsvec, mstartsvec, flearnsvec, mlearnsvec)
colnames(search_grid) <- c("fstart", "mstart", "f_learn", "m_learn")
```


## Questions
There are three major questions regarding our evaluation of `DISCent`: 

1. Model Accuracy: Can we recapture expected demographic/popgen patterns based on simulations
2. Realism: Are outputted values of $f$ and $m$ realistic
3. Model Selection: Can we use cost to distinguish the best model/data input



```{r}
#............................................................
# Part 0: Read in pieces and results 
#...........................................................
# read in model framework
migmatdf <- readRDS("simdata/00-simulation_setup/inputs/migmat_framework.RDS")
coordgrid <- readRDS("simdata/00-simulation_setup/inputs/squarecoords.rds")
# read in godisc results
discfn <- list.files("simdata/disc_results", pattern = ".RDS", full.names = T)
discrets <- lapply(discfn, function(x){
  df <- readRDS(x)
  nm <- stringr::str_replace(basename(x), ".RDS", "")
  df <- df %>% 
    dplyr::mutate(mod = nm,
                  modbase = stringr::str_split_fixed(nm, "_", n = 3)[,1],
                  modbase = forcats::as_factor(modbase),
                  rep = stringr::str_split_fixed(nm, "_", n = 3)[,2],
                  rep = forcats::as_factor(rep)) %>% 
    dplyr::relocate(mod, modbase)
  return(df)
}) %>% 
  dplyr::bind_rows()

#......................
# get cost, fs, and ms 
#......................
discrets <- discrets %>% 
  dplyr::mutate(
    final_cost = purrr::map_dbl(discret, extract_final_cost),
    final_fs = purrr::map(discret, get_fs),
    final_m = purrr::map_dbl(discret, get_ms)
  )

```


## **Q1: Model Accuracy**

### Spectrum of Costs across Models
Remember, we are ignoring reps here which represent different realizations of the `polySimIBD` model with the same initial start frameworks (e.g. lattice, etc.). For the purposes of clarification, we are going to call simulation reps, realizations. 
```{r}

discrets %>% 
  ggplot() + 
  geom_boxplot(aes(x = modbase, y = final_cost)) +
  theme_linedraw() +
  ggtitle("Costs over All Model Frameworks") + 
  theme(plot.title = element_text(hjust = 0.5))

```


### Influence of Start Params 
Remember, here we are  looking at all 135 start parameters results: 135*100*5 (135 start param combinations, 100 realizations, 5 different models). Remember, for each result there is 1 global M estimate and 25 Fi estimates (per deme). 

```{r allstart}

discrets %>% 
  dplyr::mutate(
    start_param_realization_fct = factor(start_param_realization, levels = 1:135, ordered = T)
  ) %>% 
  ggplot() + 
  geom_boxplot(aes(x = start_param_realization_fct,
                   y = final_cost)) + 
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() + 
  xlab("Start Params: 135 Unique Combinations from Above F and M starts") +
  ylab("Final Cost") + 
  ggtitle("Spectrum of Final Costs Across Different Start Parameter Runs") + 
  theme(axis.text.x = element_text(angle = 90, size = 5,
                                   vjust = 0.5, hjust = 1),
        plot.title = element_text(hjust = 0.5))




discrets %>% 
  dplyr::mutate(
    start_param_realization_fct = factor(start_param_realization, levels = 1:135, ordered = T)
  ) %>% 
  ggplot() + 
  geom_boxplot(aes(x = start_param_realization_fct,
                   y = final_m)) + 
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() + 
  xlab("Start Params: 135 Unique Combinations from Above F and M starts") +
  ylab("Final M") + 
  ggtitle("Spectrum of Final M Results Across Different Start Parameter Runs") + 
  theme(axis.text.x = element_text(angle = 90, size = 5,
                                   vjust = 0.5, hjust = 1),
        plot.title = element_text(hjust = 0.5))




discrets %>% 
  dplyr::mutate(
    start_param_realization_fct = factor(start_param_realization, levels = 1:135, ordered = T),
    finalf10 = purrr::map_dbl(final_fs, function(x) x[10,2])
  ) %>% 
  ggplot() + 
  geom_boxplot(aes(x = start_param_realization_fct,
                   y = finalf10)) + 
  theme_linedraw() + 
  facet_grid(rows = vars(modbase), scales = "free") +
  xlab("Start Params: 135 Unique Combinations from Above F and M starts") +
  ylab("Final F10") + 
  ggtitle("Spectrum of Final F10 Results Across Different Start Parameter Runs") + 
  theme(axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5))




```
**Results suggest that start parameters are important and that the user will have try several different combinations to converge on the lowest cost and best estimates**. 


### Influence of `polySimIBD` Realizations/Stochasticity on Costs and Parameter Estimates  
```{r polysimreps}

discrets %>% 
  ggplot() + 
  geom_boxplot(aes(x = rep, y = final_cost)) +
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() +
  ggtitle("Costs over All Model Frameworks w/r/t Realizations") +
  xlab("polySimIBD Realization") + 
  theme(axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5))

```
Although the above look like a `geom_point` plot, they are actually boxplots, which emphasizes that costs varied considerably w/r/t different `polySimIBD` realizations within the same framework (i.e. same migration matrix). 

```{r}

p1 <- discrets %>% 
  dplyr::mutate(
    rep_fct = factor(rep, levels = paste0("rep", 1:100), ordered = T)
  ) %>% 
  ggplot() + 
  geom_boxplot(aes(x = rep_fct,
                   y = final_m)) + 
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() + 
  xlab("polySimIBD Unique Realizations") +
  ylab("Final M") + 
  ggtitle("Spectrum of Final M Results Across Different polySimIBD Realizations \n Considering ALL Costs") + 
  theme(axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5))



p2 <- discrets %>% 
  dplyr::group_by(modbase, rep) %>% 
  dplyr::arrange(final_cost) %>%
  dplyr::slice_min(final_cost, n = 10) %>% 
  dplyr::mutate(rep_fct = factor(rep, levels = paste0("rep", 1:100), ordered = T)) %>% 
  ggplot() + 
  geom_boxplot(aes(x = rep_fct,
                   y = final_m)) + 
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() + 
  xlab("polySimIBD Unique Realizations") +
  ylab("Final M") + 
  ggtitle("Spectrum of Final M Results Across Different polySimIBD Realizations \n Considering 10 BEST Costs") + 
  theme(axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5))

#......................
# bring together
#......................
cowplot::plot_grid(p1, p2, nrow = 2)

```
**Although there is less variation in the Final Ms when it comes to those model runs w/ the start params that produced the lowest costs, still considerable variation in Final Ms across same `polySimIBD` simulation frameworks (i.e. same starting migration matrix)**.


```{r polysimrepsparamM}

p1 <- discrets %>% 
  dplyr::mutate(
    rep_fct = factor(rep, levels = paste0("rep", 1:100), ordered = T),
    finalf10 = purrr::map_dbl(final_fs, function(x) x[10,2])
  ) %>% 
  ggplot() + 
  geom_boxplot(aes(x = rep_fct,
                   y = finalf10)) + 
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() + 
  xlab("polySimIBD Unique Realizations") +
  ylab("Final F10") + 
  labs(title = "Spectrum of Final F10 Results Across Different polySimIBD Realizations \n Considering ALL Costs",
       caption = "F10 randomly selected among the 25 F (deme) options") + 
  theme(axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5))



p2 <- discrets %>% 
  dplyr::group_by(modbase, rep) %>% 
  dplyr::arrange(final_cost) %>%
  dplyr::slice_min(final_cost, n = 10) %>% 
  dplyr::mutate(rep_fct = factor(rep, levels = paste0("rep", 1:100), ordered = T),
                finalf10 = purrr::map_dbl(final_fs, function(x) x[10,2])) %>% 
  ggplot() + 
  geom_boxplot(aes(x = rep_fct,
                   y = finalf10)) + 
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() + 
  xlab("polySimIBD Unique Realizations") +
  ylab("Final F10") + 
  labs(title = "Spectrum of Final F10 Results Across Different polySimIBD Realizations \n Considering 10 Best Costs",
       caption = "F10 randomly selected among the 25 F (deme) options") + 
  theme(axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5))

#......................
# bring together
#......................
cowplot::plot_grid(p1, p2, nrow = 2)

```
**Interestingly, less variation in the Final F10 (randomly selected) when it comes to both ALL costs and BEST 10 costs - suggesting that F converges much faster than M and is not driving differences as much. This is particulalry notable given the considerable variation in Final F  across same `polySimIBD` simulation frameworks (i.e. same starting migration matrix)**.

\n
\n

**Overall, these results suggest that `polySimIBD` introduces a fair amount of stochasticity across realizations (as expected), which the model seems to adapt to - will need to look at averages to see if expected demographic/popgen patterns are recapitulated**. 


### Best Parameter Estimates by Realization
Here we have subsetted to the DISC results that produced the lowest cost from the 135 start parameter estimates by model by realization. As a result, from the 135*100*5 overall results (start params * `polySimIBD` realizations * `polySimIBD` model frameworks [migration matrices]), we have subsetted to 1*100*5 results (i.e. lowest cost across start params.) best parameter estimates by model. We expect the "average" DISC results across the `polySimIBD` simulation frameworks to match our popgen/demograph expectations (i.e. lattice, torus): in other words, expect some consistency in Fs and Ms in these results.




#### Consistency of Start Params
Here, we will look at the variation by realization of which start paramater set produced the lowest cost with respect to `polySimIBD` model frameworks (i.e. migration matrices).
```{r best1}
discretsbest <- discrets %>% 
  dplyr::group_by(modbase, rep) %>% 
  dplyr::arrange(final_cost) %>%
  dplyr::slice_min(final_cost, n = 1) %>% 
  dplyr::ungroup(rep) # one bad IBD model has a duplicated final_cost 

```
```{r}
discretsbest %>% 
  dplyr::mutate(
    start_param_realization_fct = factor(start_param_realization, levels = 1:135, ordered = T)
  ) %>% 
  ggplot() + 
  geom_boxplot(aes(x = start_param_realization_fct,
                   y = final_cost)) + 
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() + 
  xlab("Start Params: 135 Unique Combinations from Above F and M starts") +
  ylab("Final Cost") + 
  ggtitle("Spectrum of Start Parameters that Produced the Final BEST Costs Across Different Sim Frameworks") + 
  theme(axis.text.x = element_text(angle = 90, size = 5,
                                   vjust = 0.5, hjust = 1),
        plot.title = element_text(hjust = 0.5))

# manual 
# what is set 45
set45 <- unique(discretsbest$start_params[discretsbest$start_param_realization == 45])[[1]]
set45 <- unique(unname(set45))

```
**Overall, it looks like paramater set 45 produced the best results but as the figure demonstrates, this was not universally true. Set 45 was: ** `r print(set45)`. 

#### Evaluating Parameter Consistency 
```{r best1plotsF}
discretsbest %>% 
  tidyr::unnest(cols = "final_fs") %>% 
  dplyr::mutate(Deme = factor(Deme, levels = 1:25, ordered = T)) %>% 
  ggplot() + 
  geom_boxplot(aes(x = Deme, y = Final_Fis), alpha = 0.5) +
  theme_linedraw() +
  facet_grid(rows = vars(modbase), scales = "free") +
  scale_color_viridis_c() +
  ggtitle("Across Realizations (best), Demes vs Final Fs") +
  theme(axis.text.x = element_text(angle = 90),
        plot.title = element_text(hjust = 0.5))
```

^ Interesting variation in the lattice model. 


```{r best1plotsM}
discretsbest %>% 
  tidyr::unnest(cols = "final_m") %>% 
  ggplot() + 
  geom_boxplot(aes(x = modbase, y = final_m), alpha = 0.5) +
  theme_linedraw() +
  ggtitle("Across Realizations (best), Final Ms") +
  theme(axis.text.x = element_text(angle = 90),
        plot.title = element_text(hjust = 0.5))
```
^ Interesting to see that the model as no idea what to do w/ the Bad IBD - that's promising, as the distance has a good bit of error in it. 



#### Best Parameter Estimates Over Space

```{r best1plotsSPACE}
# bring together w/ spatial data
coordgrid$deme <- as.character(coordgrid$deme )
discretsum <- discretsbest %>% 
  tidyr::unnest(cols = "final_fs") %>% 
  dplyr::group_by(modbase, Deme) %>% 
  dplyr::summarise(
    n = n(),
    meanFi = mean(Final_Fis),
    varFi = var(Final_Fis),
    minFi = min(Final_Fis),
    maxFi = max(Final_Fis)
  ) %>% 
  dplyr::rename(deme = Deme) %>% 
  dplyr::left_join(., coordgrid, by = "deme") 

# need to split up b/c have different DISC scales
plotmn <- function(discretsumitem, colnm, modnm) {
  colnm <- enquo(colnm)
  plotObj <- discretsumitem %>% 
    ggplot() + 
    geom_point(aes_string(x = "longnum", y = "latnum", 
                          color = quo_name(colnm)), size = 5) + 
    geom_text(aes(x = longnum, y = latnum, label = deme),
              color = "#ffffff", size = 3) + 
    scale_color_viridis_c(paste(quo_name(colnm), " DISCs")) +
    ylim(c(-2, 24)) + xlim(c(-2, 24)) +
    ggtitle(paste(modnm, " Model")) +
    theme_linedraw() +
    theme(axis.text = element_blank(),
        plot.title = element_text(hjust = 0.5))
  return(plotObj)
}


# split out and expand to match lvls needed to run
discretsumbymod <- split(discretsum, discretsum$modbase)
discretsumbymod <- append(discretsumbymod, discretsumbymod)
discretsumbymod <- append(discretsumbymod, discretsumbymod)
discretsumbymod <- discretsumbymod[order(names(discretsumbymod))]
modnms <- names(discretsumbymod)
lvls <- rep(c("meanFi", "varFi", "minFi", "maxFi"), 5)
plotObjs <- mapply(plotmn, 
                   discretsumitem = discretsumbymod,
                   colnm = lvls, 
                   modnm = modnms, SIMPLIFY = F)
```

#### Bad Isolation by Distance
```{r}
cowplot::plot_grid(plotlist = plotObjs[1:4]) 

```

#### Good Isolation by Distance
```{r}
cowplot::plot_grid(plotlist = plotObjs[5:8]) 

```

#### Lattice
```{r}
cowplot::plot_grid(plotlist = plotObjs[9:12]) 
```

#### NeVary
```{r}
cowplot::plot_grid(plotlist = plotObjs[13:16]) 
```

#### Torus
```{r}
cowplot::plot_grid(plotlist = plotObjs[17:20]) 
```

##### Final Ms
Plot of Final M distribution. 
```{r}
# one call instead of 4
discretsbest %>% 
  tidyr::unnest(cols = "final_m") %>% 
  ggplot() + 
  geom_boxplot(aes(x = modbase, y = final_m)) +
  theme_linedraw() +
  ggtitle("Final Ms over Model Frameworks")


discretsbest %>% 
  tidyr::unnest(cols = "final_m") %>% 
  ggplot() + 
  geom_point(aes(x = final_m, y = final_cost, color = modbase)) +
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() +
  ggtitle("Final Ms vs Cost over Model Frameworks") +
  theme(plot.title = element_text(hjust = 0.5))
```

Versus M Summary Statistics 
```{r}

# table 
discretsbest %>% 
  tidyr::unnest(cols = "final_m") %>% 
  dplyr::group_by(modbase) %>% 
  dplyr::summarise(
    n = n(),
    meanM = mean(final_m),
    varM = var(final_m),
    minM = min(final_m),
    maxM = max(final_m)) %>% 
  dplyr::mutate_if(is.numeric, round, 2) %>% 
  DT::datatable(.,
                rownames = F,
                extensions='Buttons',
                options = list(
                  searching = T,
                  pageLength = 5,
                  dom = 'Bfrtip',
                  buttons = c('csv')))

```


## **Q2: Model Realism**
Because we normalized our geo-distances, they are on a scale [0,1], which means that the higher the value of M, the closer the migration term, $e^{\frac{-d}{M}}$, goes to 1. As a result, the higher the migration rate, the more panmictic a population becomes, where distances from deme to deme become neglible. **Interesting, this is the behavior that is occuring with the _lattice_ and the _torus_ models, which is consistent with the migration matrix (can only move to neighbors [left,right - no diagonals]), as it smooths distance/makes it an essentially uniform migration rate and the resulting relatedness are affected more by potential moves (i.e. beginning and ends are bounded in lattice vs no bounds in torus).** In contrast, the IBDist and NeVary models are both parameterized with migration matrices that take into account distance (as seen below: migration rate is $\frac{1}{dist}$).

```{r}

migmatplot <- migmatdf %>% 
  dplyr::mutate(longmigmat = purrr::map(migmat,
                                        function(x){ broom::tidy(as.dist(x)) })) %>% 
  dplyr::select(-c("migmat")) %>% 
  tidyr::unnest(cols = longmigmat) %>% 
  dplyr::rename(migrationrate = distance) %>% 
  dplyr::mutate(item1 = factor(item1, levels = 1:25, ordered = T),
                item2 = factor(item2, levels = 1:25, ordered = T)) %>% 
  ggplot() + 
  geom_tile(aes(x = item1, y = item2, 
                fill = migrationrate)) +
  facet_grid(rows = vars(modname), scales = "free") +
  theme_minimal() +
  xlab("Deme1") + ylab("Deme2") + 
  scale_fill_viridis_c() +
  labs(title = "Migration Rates by Deme",
       caption = "NeVary uses IBDist MigMat but varies Ne (vs constant in IBDist)") +
  theme(axis.text.x = element_text(angle = 90, size = 5),
        plot.title = element_text(hjust = 0.5),
        legend.position = "bottom")

coordmap <- coordgrid %>% 
  ggplot() + 
  geom_point(aes(x = longnum, y = latnum),
             shape = 21, fill = NA, color = "#000000",
             size = 8) + 
  geom_text(aes(x = longnum, y = latnum, label = deme)) + 
  theme_minimal() + 
  xlab("Longnum") + ylab("Latnum") + 
  labs(title = "Deme Locations",
       caption = "Migration matrix acts on deme w/ corresponding number(s)") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  ylim(c(-2, 24)) + xlim(c(-2, 24)) 


#......................
# bring together
#......................
cowplot::plot_grid(migmatplot, coordmap, nrow = 1, rel_widths = c(0.7,0.3))

```

```{r}
cowplot::plot_grid(plotlist = plotObjs[c(1,5,9,13,17)]) 
```

### Realism when M Matters
The major questin then becomes, when M is lower, is there something fundamentally different about those `polySimIBD` realizations for the IBDist and Nevary runs. Here we will take the top and bottom ten Ms and compare the input data. 


```{r}
polySimIBDret <- readRDS("simdata/polysim_results/discdat_from_polySimIBD_maestro.RDS") %>% 
  dplyr::mutate(rep = paste0("rep", rep)) %>% 
  rename(modbase = modname)

exploreMdat <- discretsbest %>% 
  dplyr::filter(modbase %in% c("IsoByDist", "NeVary"))

exploreMdattopM <- exploreMdat %>% 
  dplyr::group_by(modbase) %>% 
  dplyr::arrange(final_m) %>%
  dplyr::slice_max(final_m, n = 10) 

exploreMdatbotM <- exploreMdat %>% 
  dplyr::group_by(modbase) %>% 
  dplyr::arrange(final_m) %>%
  dplyr::slice_min(final_m, n = 10) 
#......................
# bring together top 10 and  bottom 10 
#......................
exploreMdat <- dplyr::bind_rows(exploreMdattopM, exploreMdatbotM)

#......................
# bring in raw data 
#......................
exploreMdat <- dplyr::left_join(exploreMdat, polySimIBDret,
                                by = c("modbase", "rep"))

#......................
# look at average IBD
#......................
exploreMdat <- exploreMdat %>% 
  dplyr::mutate(
     minIBD = purrr::map_dbl(discdat, function(x){min(x$gendist)}),
      maxIBD = purrr::map_dbl(discdat, function(x){max(x$gendist)}),
    avgIBD = purrr::map_dbl(discdat, function(x){mean(x$gendist)}),
     varIBD = purrr::map_dbl(discdat, function(x){var(x$gendist)}),
    sibs = purrr::map_dbl(discdat, function(x){sum(x$gendist >= 0.5)}),
    IBDgeonorm = purrr::map(discdat, function(x){
      normgeo <- (x$geodist - min(x$geodist))/(max(x$geodist) - min(x$geodist))
      x$gendist * normgeo}),
    meanIBDgeonorm = purrr::map_dbl(IBDgeonorm, mean),
    sibGeo = purrr::map(discdat, function(x){
      mingeo <- min(x$geodist)
      maxgeo <- max(x$geodist)
      mapply(function(gendist,geodist){
        if(gendist >= 0.5) {
          gendist * (geodist - mingeo)/(maxgeo - mingeo)
        } else {
          0
        }
      }, gendist = x$gendist, geodis = x$geodist)
    }),
    meansibGeo = purrr::map_dbl(sibGeo, mean)
  )



exploreMdat %>% 
  ggplot() +
  geom_point(aes(x = final_m, y = avgIBD)) +
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() + 
  xlab("Final M") + ylab("Avg IBD")

exploreMdat %>% 
  ggplot() +
  geom_point(aes(x = final_m, y = varIBD)) +
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() + 
  xlab("Final M") + ylab("Var IBD")

exploreMdat %>% 
  ggplot() +
  geom_point(aes(x = final_m, y = meanIBDgeonorm)) +
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() + 
  xlab("Final M") +
  ylab("IBD Scaled by Normalized Geodist")

exploreMdat %>% 
  ggplot() +
  geom_point(aes(x = final_m, y = sibs)) +
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() + 
  xlab("Final M") + ylab("Sib. N")



exploreMdat %>% 
  ggplot() +
  geom_point(aes(x = final_m, y = meansibGeo)) +
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() + 
  xlab("Final M") + ylab("Sib-Rate Scaled by Normalized Geodist")


```
**No obvious reason from IBD why M is lower in the bottom 10 cases versus the top 10 cases...**. Let's look to see if the costs were _that_ different for the other start param results. 

```{r}
mintreps <- exploreMdat %>% 
  dplyr::ungroup() %>% 
  dplyr::filter(final_m < 400) %>% 
  dplyr::select(c("modbase", "rep")) %>% 
  dplyr::filter(!duplicated(.))

mintreps <- discrets %>% 
  dplyr::inner_join(., mintreps, by = c("modbase", "rep")) %>% 
  dplyr::group_by(mod) %>% 
   dplyr::slice_min(final_cost, n = 10) 

mintreps %>% 
  ggplot() + 
  geom_point(aes(x = final_m, y = final_cost,
                 color = mod)) +
  facet_grid(rows = vars(modbase), scales = "free") +
  theme_linedraw() +
  theme(legend.position = "none")


```
Costs don't seem to vary greatly for the lowest and highest M, even when the M value stays constant or shifts. **Let's look at the IBD patterns for the highest and lowest final Ms in the set**: 

```{r}
#.....................
# make coord copies
#......................
c1 <- coordgrid %>% 
  dplyr::mutate(deme1 = as.character(deme))
c2 <- coordgrid %>% 
  dplyr::mutate(deme2 = as.character(deme))

#.....................
# subset to LOW M
#......................
lowm <- polySimIBDret %>% 
  dplyr::filter(modbase == "IsoByDist" & rep == "rep95")
lowmdf <- lowm$discdat[[1]]

# bring in coords
lowmdfsp <- lowmdf %>% 
  dplyr::left_join(., c1, by = "deme1") %>% 
  dplyr::left_join(., c2, by = "deme2")

# plot out
p1 <- lowmdfsp %>% 
  dplyr::filter(gendist > 0.25) %>% 
  ggplot() + 
  geom_point(data = coordgrid, 
             aes(x = longnum, y = latnum),
             color = "#000000") +
  geom_segment(aes(
    x = longnum.x, xend = longnum.y,
    y = latnum.x, yend = latnum.y, 
    color = gendist), alpha = 0.5) +
  scale_color_viridis_c() + 
  ggtitle("Lowest M for IsoByDist") + 
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "bottom")

#.....................
# subset to HIGH M
#......................
highm <- polySimIBDret %>% 
  dplyr::filter(modbase == "IsoByDist" & rep == "rep67")
highmdf <- highm$discdat[[1]]

# bring in coords
highmdfsp <- highmdf %>% 
  dplyr::left_join(., c1, by = "deme1") %>% 
  dplyr::left_join(., c2, by = "deme2")

# plot out
p2 <- highmdfsp %>% 
  dplyr::filter(gendist > 0.25) %>% 
  ggplot() + 
  geom_point(data = coordgrid, 
             aes(x = longnum, y = latnum),
             color = "#000000") +
  geom_segment(aes(
    x = longnum.x, xend = longnum.y,
    y = latnum.x, yend = latnum.y, 
    color = gendist), alpha = 0.5) +
  scale_color_viridis_c() + 
  ggtitle("Highest M for IsoByDist") + 
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "bottom")

#......................
# bring together
#......................
cowplot::plot_grid(p1, p2, nrow = 1)

```
**This shows that the higher "node" density of the IBD connections is likely driving the lower M result, which is exactly what we hope. It is interesting that there is some kind of inflection point in the ?saturation? or IBD that makes M more important. However, of all the above signals, this is the most clear that M is actually picking up an important pattern of IBD/inbreeding**. 



## **Q3: Model Selection**
Given that reps here match between bad-IBD and good-IBD, can do a paired cost difference. 
```{r}
discretsbest_badIBD <- discretsbest %>% 
  dplyr::ungroup() %>% 
  dplyr::filter(modbase == "badIsoByDist") 


discretsbest_goodIBD <- discretsbest %>% 
  dplyr::ungroup() %>% 
  dplyr::filter(modbase == "IsoByDist") 

comparebadgood <- dplyr::inner_join(discretsbest_badIBD, discretsbest_goodIBD, by = "rep") %>% 
  dplyr::mutate(
    costdiff = final_cost.x - final_cost.y
  )


comparebadgood %>% 
  ggplot() + 
  geom_boxplot(aes(y = costdiff)) +
  theme_linedraw() + 
  xlab("Cost Diff.") + 
  ggtitle("Final Costs of Bad IsoByDist (wrong Distances) vs Correct Distance IsoByDist", subtitle = "Eq is $Bad - Good$, so negative numbers indicate cost identified good as better model (minimizing cost)") +
  theme(axis.text.x = element_blank())



comparebadgood %>% 
  ggplot() + 
  geom_point(aes(x = final_m.x, y = final_m.y, 
                 color = costdiff)) +
  theme_linedraw() + 
  scale_color_viridis_c() + 
  xlab("Final M Bad IsoByDist") + 
  ylab("Final M IsoByDist") + 
  ggtitle("Final Costs of Bad IsoByDist (wrong Distances) vs Correct Distance IsoByDist", subtitle = "Eq is $Bad - Good$, so negative numbers indicate cost identified good as better model (minimizing cost)") 


```

